\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-01-20 src/tex/ms.tex

% Abstract with filler text
\begin{abstract}
    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.

@@

The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
\AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
% \AZ{Add PPE and maybe TIGER refs..?}
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
All of these tasks require significant computation.
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.

Derivatives are ubiquitously useful throughout data analysis tasks.
For instance, during PE, gradients are used to improve performance of the samplers 
% \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
(e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

@@

For maximum utility, a waveform needs to be fast to evaluate.
Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple. \dfm{I'd soften this statement a little bit. In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster! Where are these benchmarks implemented? Would it be easy for me to try them out?}
Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
}
Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
Waveform calls can therefore be evaluated in parallel, massively improving performance.
Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

Forecasting the sensitivity of future experiments is an essential task in GW science.
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

@@

Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call \AZ{used by \texttt{GWbench}}.
As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.

@@

In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}. 
We therefore expect \jax waveforms to be \sout{crucial} \AZ{highly beneficial} to future PE efforts in GW astronomy, especially for low-latency pipelines.

>>>>

% Abstract with filler text
\begin{abstract}
    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
    \kw{I would cut everything before out.}
    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.

@@

The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
\kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
\AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
% \AZ{Add PPE and maybe TIGER refs..?}
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
\kw{Need more reference and some rewording}
All of these tasks require significant computation.
\kw{This could be triggering to some people, since population analysis could be quite cheap}
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.

Derivatives are ubiquitously useful throughout data analysis tasks.
For instance, during PE, gradients are used to improve performance of the samplers 
% \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
(e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
\kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

@@

For maximum utility, a waveform needs to be fast to evaluate.
Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple. \dfm{I'd soften this statement a little bit. In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster! Where are these benchmarks implemented? Would it be easy for me to try them out?}
Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
}
Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
Waveform calls can therefore be evaluated in parallel, massively improving performance.
Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
\kw{We should add some more detail related to this comparison. We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part). The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.}
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

@@

Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call \AZ{used by \texttt{GWbench}}.
\kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.

@@

In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}. 
We therefore expect \jax waveforms to be \sout{crucial} \AZ{highly beneficial} to future PE efforts in GW astronomy, especially for low-latency pipelines.
\kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
